---
title: "Check_In"
output:
  pdf_document: default
  html_document: default
date: "2025-09-25"
---

```{r}
# set working directory 
setwd("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/")

# 1.a.i read in data 
counts <- read.csv("data/counts.csv", row.names = 1, check.names = FALSE, stringsAsFactors = FALSE)
meta_data <- read.csv("data/meta_data.csv")
meta_data <- data.frame(meta_data)
```
For Gene 1: I selected the gene ENSG00000003402.20 (CFLAR) because it is 
implicated in apoptosis and immune response regulation. In cancer, elevated 
expression of CFLAR is correlated with resistance to therapy 
(Zhang et al., J Inflamm Res, 2025)

For Gene 2: I selected the gene ENSG00000000003.15 (TSPAN6) because it is 
involved in extracellular signalling to recruit immune cells the tumor 
microenvionrment (Molostvov et al., Cell Rep, 2023)

```{r}
# 2.a.  
# prep for figure generation 
library(ggplot2)

# make data frame with both genes
gene_data <- data.frame(
  barcode = colnames(counts),
  gene1 = as.numeric(counts['ENSG00000003402.20', ]),
  gene2 = as.numeric(counts['ENSG00000000003.15', ]))

# pull in Hunter's Cooler code 
coolor_code <- cool_palette <- c(
  "#E3E9EA",
  "#B8C4C2",
  "#7A9E9F",
  "#527C83",
  "#314D58",
  "#585858"
)

# set themes for subsequent plots
theme_set(
  theme_minimal(base_family = "Helvetica", base_size = 12) +
    theme(
      # backgrounds
      plot.background  = element_rect(fill = coolor_code[1], color = NA),
      panel.background = element_rect(fill = coolor_code[1], color = NA),

      # grid lines
      panel.grid.major = element_line(color = coolor_code[2], linewidth = 0.4),
      panel.grid.minor = element_line(color = coolor_code[2], linewidth = 0.2, linetype = "dotted"),

      # text and titles
      plot.title = element_text(
        hjust = 0.5,
        face = "bold",
        size = 14,
        color = coolor_code[5]
      ),
      axis.title = element_text(face = "bold", color = coolor_code[5]),
      axis.text  = element_text(color = coolor_code[5]),

    )
)

```


```{r}
# 3.b. Generate summary statistics 
print(summary(gene_data$gene1))
print(sd(gene_data$gene1))

print(summary(gene_data$gene2))
print(sd(gene_data$gene2))
```


```{r}
# 3. Visualization
# 3.a.- Histogram
p <- ggplot(gene_data, aes(x = gene1)) +
  geom_histogram(aes(y = after_stat(count)), binwidth = 500, 
                 fill = coolor_code[3], color = coolor_code[5]) +
  geom_density(aes(y = after_stat(count) * 500), color = coolor_code[5], 
               linewidth = 1, alpha = .5) +
  labs(
    title = "Histogram of CFLAR Expression Values",
    x = "Expression Values",
    y = "Frequency"
  )

print(p)
png("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/figures_tables/hist.png", width = 2000, height = 1500, res = 300)
print(p)
dev.off()
```


```{r}
# 3.b. - Scatter plot
library(ggpubr)


# plot gene1 against gene2
p <- ggplot(gene_data, aes(x = gene1, y = gene2)) +
  # add scatter plot 
  geom_point(
    fill  = coolor_code[3],
    color = coolor_code[5],
    shape = 21,
    size  = 4,
    stroke = .3,             
    alpha = 0.65
  ) +
  # add regression line with cooler code colors
  geom_smooth(
    method = "lm",
    se = TRUE, # add confidence interval 
    color = coolor_code[6],   
    fill  = coolor_code[6], 
    alpha = 0.7,
    linewidth = 1.1
  ) +
  labs(
    title = "CFLAR | TSPAN6",
    x = "CFLAR",
    y = "TSPAN6") +
  # add regression line equation
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), sep = "~~~")),
    label.x.npc = "left",   # x position. - leftcorner 
    label.y.npc = 0.95,      # y position - upper corner
    color = coolor_code[6],
    size = 4
  ) +
  # add correlation coefficient and p value 
  stat_cor(
    aes(label = paste(after_stat(r.label), after_stat(p.label), sep = "~`,`~")),
    label.x.npc = "left",    # same x position 
    label.y.npc = 0.87,       # slightly below other y 
    color = coolor_code[6],
    size = 4
  )

print(p)
png("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/figures_tables/scatter.png", width = 2000, height = 1500, res = 300)
print(p)
dev.off()

```




```{r}
# 3.c. - Violin plot
# define data frame for gene of interest
meta_col <- "tissue_type"   # <-- change this to any column in meta (e.g., "definition")

# merge gene and meta data 
gene_meta <- merge(gene_data[, c("barcode", "gene1")], meta_data[, c("barcode", meta_col)], by = "barcode", all.x = TRUE)

# make sure tissue_type is a factor not a continuous variable
gene_meta$tissue_type <- as.factor(gene_meta$tissue_type)

library(ggplot2)
# make violin plot
p <- ggplot(gene_meta, aes(x = tissue_type, y = gene1, fill = tissue_type)) +
  geom_violin() + 
  scale_fill_manual(values = coolor_code[2:5]) +
  labs(
    x = "Tissue Type",
    y = "CFLAR Gene Expression",
    fill = "Tissue Type", # reformat fill legend
    title = "CFLAR Expression Distribution by Tissue Type"
  )
print(p)
png("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/figures_tables/violin.png", width = 2000, height = 1500, res = 300)
print(p)
dev.off()

```
Heatmap Analysis: 
Select 10 genes: Choose a set of 10 different genes from the count matrix for your heatmap.
Generate a Heatmap: Use the ComplexHeatmap package in R to create a heatmap of the count data for the selected genes. 
Add an Annotation Bar: Include an annotation bar reflecting your chosen covariate for further context and interpretation of the data.


```{r}
# implement Chayce's method for changing ensemble ID ==> Gene Names 
library(biomaRt)

# find top ten most variable genes
gene_vars <- apply(counts, 1, var)
genes_to_plot <- names(sort(gene_vars, decreasing = TRUE)[1:10])

# subset genes 
counts_subset <- counts[genes_to_plot, ]

# Remove version suffix from ensembl IDs
genes_to_plot <- sub("\\.\\d+$", "", genes_to_plot)
rownames(counts_subset) <- sub("\\.\\d+$", "", rownames(counts_subset))


# Connect to ensembl with mirror
ensembl_df <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl", # human genes
  mirror = "useast"   # use mirror because original ensembl not available
)

# get ensembl gene names for genes_to_plot values 
gene_names <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters   = "ensembl_gene_id",
  values    = genes_to_plot,
  mart      = ensembl_df
)



# make rownames ensembl_gene_id
rownames(gene_names) <- gene_names$ensembl_gene_id

# remove ensembl_gene_id
gene_names$ensembl_gene_id <- NULL

# check 
head(gene_names)

#merge external names with counts_subset
subset_merged <- merge(gene_names, counts_subset,  by = "row.names", all = TRUE)

# rinse and repeat 
# make rownames external_gene_name
rownames(subset_merged) <- subset_merged$external_gene_name

# remove ensembl_gene_id
subset_merged$external_gene_name <- NULL
subset_merged$Row.names <- NULL


head(subset_merged)
```


```{r}
# Heatmap Analysis
library (ComplexHeatmap)

## Heatmap annotations
column_ha <- HeatmapAnnotation(
  # change annotation track names 
  "Tissue Type" = meta_data$tissue_type,
  "Gender"      = meta_data$gender,
  # set colors 
  col = list(
    "Tissue Type" = c(Tumor = "red", Normal = "blue"),
    "Gender"      = c(male  = "green", female = "purple")
  ),
  # change legend titles and labels 
  annotation_legend_param = list(
    "Tissue Type" = list(title = "Tissue Type", labels = c("Normal Tissue","Tumor Tissue")),
    "Gender"      = list(title = "Gender",      labels = c("Female","Male"))
  ),
  # position annotations on the right 
  annotation_name_side = "right",   
  show_annotation_name = TRUE
)


# Create heatmap
p <- Heatmap(
  subset_merged,
  name = "Counts",
  top_annotation = column_ha,
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_columns = TRUE,
  cluster_rows = TRUE,
  column_title = "Samples",
  row_title = "Genes",
  heatmap_legend_param = list(title = "Count")
)

print(p)
png("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/figures_tables/heatmap.png", width = 2000, height = 1500, res = 300)
print(p)
dev.off()
```

```{r}
# make a new plot! I chose a cumulative distribution function 
# Cumulative Distribution Function
p <- ggplot(gene_data, aes(x = gene1)) +
  # shaded area under ECDF
  stat_ecdf(
    geom = "area",
    fill = coolor_code[3],
    alpha = 0.25
  ) +
  # ECDF line 
  stat_ecdf(
    geom = "step",
    color = coolor_code[3],
    linewidth = 1
  ) +
  # median line
  geom_vline(
    xintercept = median(gene_data$gene1, na.rm = TRUE),
    linetype = "dashed",
    color = coolor_code[5],
    linewidth = 0.6
  ) +
  # add median line text
  annotate("text", x =5600 , y = 0.75, label = "Median",
           vjust = -0.6, color = coolor_code[5], size = 3) +
  labs(
    title = "CDF of CFLAR",
    x = "Expression Values",
    y = "Cumulative Probability"
  ) 

print(p)
png("/Users/jmelb/Projects/HDS-Foundations_of_Data_Science/Final_Project/figures_tables/CDF.png", width = 2000, height = 1500, res = 300)
print(p)
dev.off()



```



